Multiple Dice Working as One: CAD Flows and Routing Architectures for Silicon Interposer FPGAs

### 2014-12-30 doc/journal/fpgapaperfinal/FPGAPaper/fpga2014paper.tex
Ehsan's final revisions. removed a couple of figures and wordsmith the rest of my sections. I should re-read the placer optimizations section to make sure the formulas are consistent.
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4569 8e3573b8-cf2c-4f14-ef6d-137439e28b8b

\begin{enumerate}
\item Fan-in Transfer for Interposer Wires\\
For an architecture with \textit{\% wires cut} $=0$, all vertical routing wires have a corresponding interposer wire that connects them to the adjacent die. Practical interposers, however, can only accommodate a fraction of all vertical routing wires. When \textit{fanin transfer} option is \textit{on}, we place a multiplexer at the input of each available interposer wire and connect some adjacent ``dangling" vertical routing wires to the multiplexer. This allows for additional flexibility in routing by ensuring that \textit{every} vertical routing wire could \textit{potentially} drive an interposer wire through a multiplexer. As an example, if \textit{\% wires cut} $=75\%$, 3 of every 4 interposer wires are cut. Therefore, we place a 4-to-1 multiplexer at the input of the one interposer wire that is not cut, and we connect the 3 vertical routing wires that do not have a corresponding interposer wire to the multiplexer.

\item Fan-out Transfer for Interposer Wires\\
When \textit{fanout transfer} option is \textit{on}, we allow the interposer wire output to drive not only its own corresponding vertical wire at the other side of the cut, but also the other ``dangling" vertical routing wires that are not driven by anything because their corresponding interposer wires are cut.

\item Bidirectional Interposer Wires\\
Dominant routing architectures in modern FPGAs use unidirectional wires~\cite{unidirectional, lewis2005stratix}. Such wires can only be driven at one point, normally their start point. When this option is \textit{on}, we allow interposer wires to be driven from one end or the other by adding tri-state buffers at both ends.
\end{enumerate}

@@

For every vertical routing wire in the channel, we add a new routing node (``\textit{interposer node}") into the rrgraph. Then, we determine rrnodes (e.g. vertical routing wires) that cross a cutline in the rrgraph, and we replace them with 2 rrnodes: one that is above the cutline and one that is below the cutline. We properly transfer fan-ins and fan-outs of the original node onto the two newly created nodes and we connect the two nodes using an interposer node. Finally, we remove \textit{all} edges from \textit{\%wires cut} percent of interposer nodes and increase the switch delay of output edges of the remaining interposer nodes by \textit{increased delay}. Figure~\ref{fig:rrgraph_ops} shows an example of these operations.

@@

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:timing_eq_full}
Timing\_Cost=\sum_{\forall i, j \subset circuit} delay(\Delta x_{ij},\Delta y_{ij}) \times \\ 
criticality(i,j)
\end{multline}
\setlength{\arraycolsep}{5pt}

@@

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:timing_cost}
delay(i,j) = delay(\Delta x_{ij}, \Delta y_{ij}) + \\
times\_crossed(i,j) \times delay\_increase
\end{multline}
\setlength{\arraycolsep}{5pt}

where $times\_crossed(i,j)$ is the number of times this path has to cross the interposer to go from $(x_i,y_i)$ to $(x_j,y_j)$ and $delay\_increase$ is the timing penalty of crossing the interposer.

@@

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:bbcost}
wiring\_cost_{orig} = \sum_{n=1}^{N_{nets}} q(n) \times [\frac{bb_x(n)}{avg\_chanx\_W(n)} + \\
\frac{bb_y(n)}{avg\_chany\_W(n)}]
\end{multline}
\setlength{\arraycolsep}{5pt}

where $bb_x(n)$ and $bb_y(n)$ are the dimensions of the bounding box of net \textit{n} in the x and y directions, respectively. $avg\_chanx\_W$ and $avg\_chany\_W$ are the average x-directed and y-directed channel widths over this bounding box. Finally, $q(n)$ is a function obtained from \cite{icann} which models the fact that bounding boxes underpredict the required routing for high fanout nets. $q(n)$ slowly increases with the fanout of net \textit{n}.

VPR's wiring cost also considers the FPGA to be homogeneous, and uses only the number of nets, the size of the net's bounding box and the average number of wires per channel to calculate the cost. Thus, to account for the reduced connectivity near the cutlines we create an extra cost term, \textit{cut\_cost}. This new cost is added to (\ref{eq:bbcost}) to create the total wiring cost.

\setlength{\arraycolsep}{0.0em}
\begin{equation}
\label{eq:total_wiring}
wiring\_cost = wiring\_cost_{orig} + cut\_cost
\end{equation}
\setlength{\arraycolsep}{5pt}

In~\cite{interposer2014}, Pereira et al tested several different \textit{cut\_cost} formulations, as well as different weighting for each. They found that the following \textit{cut\_cost} results in the best quality of results in terms of minimum routable channel width and critical path delay.

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:cost2}
cut\_cost = \sum_{n=1}^{N_{nets}} C' \times bbHeight(i) \times times\_crossed(n)
\end{multline}

@@

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:lookahead_cost}
Total\_Cost(r) = \sum_{\forall l \in RT(i,r)} Cost(l) + \alpha \times ExpectedCost(r,j)
\end{multline}
\setlength{\arraycolsep}{5pt}

@@

In a similar way, we modify the $ExpectedCongestionCost(r,j)$ in (\ref{eq:expected_routing_cost}) to account for the new interposer routing nodes.

The interposer-aware router lookahead is able to make routing decisions more quickly given the large delay difference of a path that crosses the interposer as opposed to a path that doesn't.

@@

The new router lookahead changes the priority of the routes discovered by the router. Using the new router lookahead function increases the critical path delay by 1.3\% on average as shown in Figure~\ref{fig:lookahead_fmax}. We consider this small quality loss acceptable and use the enhanced router lookahead in all our experiments due to its greatly reduced runtime.

\begin{figure}[!hb]
\centering
\includegraphics[width=\linewidth]{router_opt_fmax.eps}
\caption{Impact of new vs. original router lookahead on the critical path delay.}
\label{fig:lookahead_fmax}
\end{figure}

@@

To analyze the impact of the number of cut wires we ran experiments varying only this parameter while leaving the other parameters constant. We use \textit{number of cuts} $= 1$ (corresponding to two dice), \textit{increased delay} $= 1ns$, \textit{bidirectional interposer wires = on}, \textit{fanin transfer = on}, and \textit{fanout transfer = on}.

@@

The critical path delay, depicted in Figure~\ref{fig:standard_crit}, on the other hand, is not strongly influenced by the percentage of wires cut, as the critical path delay at 80\% of the wires cut is essentially the same as at 0\% wires cut. Note, however, that this same critical path delay is achieved at a much higher channel width when \textit{\% wires cut} is greater than 80\%.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{standard_crit_path_new.eps}
\caption{Critical path delay vs. \textit{\% wires cut} for 2 dice and $1ns$ of \textit{delay increase}.}
\label{fig:standard_crit}
\end{figure}

@@

%   - Delay vs % wires cut & wire delay
\subsection{Circuit Speed vs. Interposer Delay}

\begin{figure}[!b]
\centering
\includegraphics[width=\linewidth]{delays_crit_path_new.eps}
\caption{Critical path delay vs. \textit{\% wires cut} for 2 dice and $0.0$, $0.5$, $1.0$ and $1.5ns$ of \textit{delay increase}.}
\label{fig:delays_crit}
\end{figure}

To investigate the impact of the interposer delay (\textit{delay increase}), we keep all other interposer parameters constant and sweep the \textit{delay increase} from 0 to 1.5 \textit{ns}. We use \textit{number of cuts} $= 1$ (i.e. two dice), \textit{bidirectional interposer wires = on}, \textit{fanin transfer = on}, and \textit{fanout transfer = on}.

@@

>>>>

\begin{enumerate}
\item Fan-in Transfer for Interposer Wires\\
For an architecture with \textit{\% wires cut} $=0$, all vertical routing wires have a corresponding interposer wire that connects them to the adjacent die. Practical interposers, however, can only accommodate a fraction of all vertical routing wires. When \textit{fanin transfer} option is \textit{on}, we place a multiplexer at the input of each available interposer wire and connect adjacent ``dangling" vertical routing wires to the multiplexer. This enables additional flexibility in routing by ensuring that \textit{every} vertical routing wire could \textit{potentially} drive an interposer wire through a multiplexer. As an example, if \textit{\% wires cut} $=75\%$, 3 of every 4 interposer wires are cut. Therefore, we place a 4-to-1 mux at the input of the one interposer wire that is not cut, and we connect the 3 vertical routing wires that do not have a corresponding interposer wire to the multiplexer.

\item Fan-out Transfer for Interposer Wires\\
When \textit{fanout transfer} option is \textit{on}, we allow an interposer wire to drive not only its own corresponding vertical wire at the other side of the cut, but also other ``dangling" vertical routing wires that are not driven by anything because their corresponding interposer wires are cut.

\item Bidirectional Interposer Wires\\
Dominant routing architectures in modern FPGAs use unidirectional wires~\cite{unidirectional, lewis2005stratix} which can only be driven at one point, normally their start point. When this option is \textit{on}, we allow interposer wires to be driven from one end or the other by adding tri-state buffers at both ends.
\end{enumerate}

@@

For every vertical routing wire in the channel, we add a new routing node (``\textit{interposer node}") into the rrgraph. Then, we determine rrnodes (e.g. vertical routing wires) that cross a cutline in the rrgraph, and we replace them with 2 rrnodes: one that is above the cutline and one that is below the cutline. We properly transfer fan-ins and fan-outs of the original node onto the two newly created nodes and connect the two nodes using an interposer node. Finally, we remove \textit{all} edges from \textit{\%wires cut} percent of interposer nodes and increase the switch delay of output edges of the remaining interposer nodes by \textit{increased delay}. Figure~\ref{fig:rrgraph_ops} shows an example of these operations.

@@

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:timing_eq_full}
Cost_{T}=\sum_{\forall i, j \subset circuit} delay(\Delta x_{ij},\Delta y_{ij}) \times 
Criticality(i,j)
\end{multline}
\setlength{\arraycolsep}{5pt}

@@

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:timing_cost}
delay(i,j) = delay(\Delta x_{ij}, \Delta y_{ij}) + \\
N_{crossed}(i,j) \times delay\_increase
\end{multline}
\setlength{\arraycolsep}{5pt}

where $N_{crossed}(i,j)$ is the number of times this path has to cross the interposer to go from $(x_i,y_i)$ to $(x_j,y_j)$ and $delay\_increase$ is the timing penalty of crossing the interposer.

@@

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:bbcost}
Cost_{W\_orig} = \sum_{n=1}^{N_{nets}} q(n) \times [\frac{bb_x(n)}{W_{chanx}} + \frac{bb_y(n)}{W_{chany}}]
\end{multline}
\setlength{\arraycolsep}{5pt}

where $bb_x(n)$ and $bb_y(n)$ are the dimensions of the bounding box of net \textit{n} in the x and y directions, respectively. $W_{chanx}$ and $W_{chany}$ are the average x-directed and y-directed channel widths over this bounding box. Finally, $q(n)$ is a function obtained from \cite{icann} which models the fact that bounding boxes underpredict the required routing for high fanout nets. $q(n)$ slowly increases with the fanout of net \textit{n}.

VPR's wiring cost also considers the FPGA to be homogeneous, and uses only the number of nets, the size of the net's bounding box and the average number of wires per channel to calculate the cost. Thus, to account for the reduced connectivity near the cutlines we create an extra cost term, $Cost_{cut}$. This new cost is added to (\ref{eq:bbcost}) to create the total wiring cost.

\setlength{\arraycolsep}{0.0em}
\begin{equation}
\label{eq:total_wiring}
Cost_{W} = Cost_{W\_orig} + Cost_{cut}
\end{equation}
\setlength{\arraycolsep}{5pt}

In~\cite{interposer2014}, Pereira et al tested several different $Cost_{cut}$ formulations, and different weighting for each. They found that the following $Cost_{cut}$ results in the best quality of results in terms of minimum routable channel width and critical path delay.

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:cost2}
Cost_{cut} = \sum_{n=1}^{N_{nets}} C' \times bbHeight(i) \times N_{crossed}(n)
\end{multline}

@@

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:lookahead_cost}
Cost_{Total}(r) = \sum_{\forall l \in RT(i,r)} Cost(l) + \alpha \times ExpectedCost(r,j)
\end{multline}
\setlength{\arraycolsep}{5pt}

@@

We modify $ExpectedCongestionCost(r,j)$ in (\ref{eq:expected_routing_cost}) in a similar way to account for the new interposer routing nodes.

The interposer-aware router lookahead makes routing decisions more quickly given the large delay difference of a path that crosses the interposer as opposed to a path that doesn't.

@@

The new router lookahead changes the priority of the routes discovered by the router. Using the new router lookahead function increases the critical path delay by 1.3\% on average. We consider this small quality loss acceptable and use the enhanced router lookahead in all our experiments due to its greatly reduced runtime.

@@

To analyze the impact of the number of cut wires we ran experiments varying only this parameter while leaving the other parameters constant. We use \textit{number of cuts} $= 1$ (i.e. two dice), \textit{increased delay} $= 1ns$, \textit{bidirectional interposer wires = on}, \textit{fanin transfer = on}, and \textit{fanout transfer = on}.

@@

@@

The critical path delay, depicted in Figure~\ref{fig:delays_crit}, on the other hand, is not strongly influenced by the percentage of wires cut, as the critical path delay at 80\% of the wires cut is essentially the same as at 0\% wires cut. Note, however, that this same critical path delay is achieved at a much higher channel width when \textit{\% wires cut} is greater than 80\%.

%   - Delay vs % wires cut & wire delay
\subsection{Circuit Speed vs. Interposer Delay}
To investigate the impact of the interposer delay (\textit{delay increase}), we keep all other interposer parameters constant and sweep the \textit{delay increase} from 0 to 1.5 \textit{ns}. We use \textit{number of cuts} $= 1$ (i.e. two dice), \textit{bidirectional interposer wires = on}, \textit{fanin transfer = on}, and \textit{fanout transfer = on}.

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{delays_crit_path_new.eps}
\caption{Critical path delay vs. \textit{\% wires cut} for 2 dice and $0.0$, $0.5$, $1.0$ and $1.5ns$ of \textit{delay increase}.}
\label{fig:delays_crit}
\end{figure}
